Applying Deep Learning to Acoustic Analysis in the PICO-60 Dark Matter Detector

### 2018-08-06 documentation/paper.tex
Reorganized introduction

\usepackage[margin=1in]{geometry}
\usepackage{textcomp}
\usepackage{parskip}
\usepackage{amsmath}
\usepackage[backend=biber, style=alphabetic, sorting=ynt]{biblatex}
\usepackage{graphicx}
\graphicspath{{./images/}}
\usepackage{wrapfig}

@@

\section{Introduction}

\subsection{Objective}

In experiments for dark matter detection, one of the most important and challenging hurdles to overcome is that of background radiation. Unwanted particles from a variety of sources can produce event signatures that appear very similar to those of dark matter candidates. In the PICO-60 bubble chamber experiment for detection of weakly interacting massive particles (WIMPs), these came in the form of neutrons (from unknown sources) and alpha particles (from decay of radon and associated daughter isotopes).

@@

\begin{enumerate}
    \item A detailed simulation of the physical environment is often used to produce an accurate discriminator. This takes a long time even for a team of experienced physicists, and will have to be updated whenever the physical variables of the experimental apparatus change.
    \item The primary intention of many dark matter experiments, which is to reduce background radiation to the lowest level possible, means that there will be a very small number of background events detected. This places a severe limit on the amount of data that is available to optimize such a discriminator.
    \item Furthermore, what little data is available to optimize a discriminator very often contains impurities, because whatever background radiation is present during WIMP detection runs is also present during calibration runs. These are difficult to separate without already having access to a functional discriminator, creating a ``chicken or the egg'' problem (from which the most common way to escape is to create a time- and labor-intensive simulation).
    \item When a discriminator is optimized on the impure training data available, for which ground truths are defined based on the calibration run of origin, it is really being optimized to discriminate between calibration runs, rather than between classes of particles. Without careful consideration, these two goals may not match up. For instance, a discriminator can undesirably fit on noise that varies between calibration runs.
\end{enumerate}

The objective of this study is to solve all 3 of these problems using machine learning. This is done using semi-supervised algorithms that simultaneously learn an effective discriminator and improves the purity of their own training data sets.

@@

\subsection{Existing Work}

During PICO-60 analysis, a function known as Acoustic Parameter was derived from simulations. It discriminates between alpha particles and nuclear recoils based on audio data (specifically the banded Fourier transform $\beta_{8}$, which is detailed in section \textbf{2.1.2}).

WIMP candidates in the cross-section and mass ranges that the PICO-60 detector is sensitive to are predicted to produce nuclear recoils with the same properties as those created by neutrons. The key difference is that neutrons frequently (but not always) scatter and produce multiple bubbles, where the extremely small predicted cross sections of WIMP candidates mean that they should almost invariably create just one bubble. This means single-bubble neutron events can be used to optimize a discriminator to detect WIMP events.

There was no explicit alpha particle calibration data collected as part of the PICO-60 experiment. However, the previous analysis revealed that there were no WIMP candidates detected. Also, based on the known ratio of multiple- to single-bubble neutron events, it is unlikely that there is more than 1 single-bubble neutron event throughout the background runs. This means they can be used as a high-purity alpha calibration set.

While these two known event sets were only used for minor calibrations of Acoustic Parameter, there is no reason they cannot be used to optimize a machine learning discriminator from scratch.

\section{Data Formats}

\subsection{Piezo-Derived}

\subsubsection{Raw Waveform}

\subsubsection{Fourier Transform}

\subsection{Camera-Derived}

\subsubsection{Image Window Sequence}

\subsubsection{3D Position}

\section{Data Cuts}

\subsection{Basic Quality Cut}

\subsection{Bubble Multiplicity Cut}

\subsection{Wall Cut}

\subsubsection{Fiducial Cut}

\subsubsection{Pressure Cut}

\subsubsection{Acoustic Cut}

>>>>

\usepackage[margin=1in]{geometry}
\usepackage{textcomp}
\usepackage{parskip}
\usepackage{amsmath}
\usepackage[backend=biber, style=alphabetic, sorting=ynt]{biblatex}
\usepackage{graphicx}
\usepackage{wrapfig}

\graphicspath{{./images/}}

@@

\section{Objective}

In experiments for dark matter detection, one of the most important and challenging hurdles to overcome is that of background radiation. Unwanted particles from a variety of sources can produce event signatures that appear very similar to those of dark matter candidates.

@@

\begin{enumerate}
    \item A detailed model of the physical environment is often used to produce an accurate discriminator. This takes a long time even for a team of experienced physicists, and will have to be updated whenever the physical variables of the experimental apparatus change.

    \item The primary intention of many dark matter experiments, which is to reduce background radiation to the lowest level possible, means that there will be a very small number of background events detected. This places a severe limit on the amount of data that is available to optimize such a discriminator.

    Furthermore, what little data is available to optimize a discriminator very often contains impurities, because whatever background radiation is present during WIMP detection runs is also present during calibration runs. These are difficult to separate without already having access to a functional discriminator, creating a ``chicken or the egg'' problem (from which one of the most common ways to escape is to create a time- and labor-intensive simulation).
\end{enumerate}

The objective of this study is to explore a potential methodology that resolves both of these problems using techniques based on machine learning. These include supervised learning algorithms that learn to separate calibration data sets, and semi-supervised learning algorithms that simultaneously learn an effective discriminator and improve the purity of their own training data sets.

\section{Previous PICO-60 Analysis}

\subsection{Background}

The PICO-60 bubble chamber experiment for detection of weakly interacting massive particles (WIMPs) was completed in 2017. It provides an excellent basis on which to develop and test the efficacy of such techniques as a potential tool for event discrimination, relative to the conventional techniques used during initial analysis of the experiment.

WIMP candidates in the cross section and mass ranges that the PICO-60 detector is sensitive to are predicted to produce nuclear recoils with the same properties as those created by neutrons. The key difference is that neutrons frequently (but not always) scatter and produce multiple bubbles, where the extremely small predicted cross sections of WIMP candidates mean that they should almost invariably create just one bubble. This means single-bubble neutron events can be used to optimize a discriminator to detect WIMP events.

The approximate expected ratio between single-bubble and multiple-bubble events generated by neutrons is known. A measured quantity of single-bubble events generated by nuclear recoils, in excess of this prediction, would be indicative of WIMP interactions. Consequently, it is imperative to isolate only those events associated with nuclear recoils.

\subsection{Discrimination Techniques}

In this study, newly developed methodologies were compared to the following discrimination techniques developed previously for the PICO-60 experiment.

\subsubsection{Acoustic Parameter}

Acoustic Parameter is the technique used for event discrimination in the original PICO-60 analysis. It is a function that discriminates between alpha particles and nuclear recoils based on audio data. It was derived based on a physical model of the bubble chamber, and was tuned using calibration runs using neutron sources to produce events in the PICO-60 apparatus.

For Acoustic Parameter to be sufficiently accurate, the audio data must first be converted into the frequency domain (specifically, with the banded Fourier transform $\beta_{8}$), and preprocessed with the position correction function $PosCor(\chi)$. In addition, a set of data cuts must be applied to remove certain classes of events. These conversion, preprocessing, and data cutting steps were re-applied during this study, and are described in section \ref{data_formats}.

\subsubsection{Machine Learning}

As part of the PICO-60 experiment, some testing was conducted to determine whether machine learning could be a viable technique to consider for a discrimination function. The results, while not definitive, indicated that the general technique warranted further investigation.

\subsection{Data Sources and Formats} \label{data_formats}

\subsubsection{Piezo-Derived}

\textit{Raw Waveform}

\textit{Fourier Transform}

\subsubsection{Camera-Derived}

\textit{Image Window Sequence}

\textit{3D Position}

\subsection{Data Cuts}

\subsubsection{Basic Quality Cut}

\subsubsection{Bubble Multiplicity Cut}

\subsubsection{Wall Cut}

\textit{Fiducial Cut}

\textit{Pressure Cut}

\textit{Acoustic Cut}

\section{Machine Learning Techniques}

@@
